Initial estuarine response to the nutrient-rich Piney Point discharge into Tampa Bay, Florida

### 2021-11-10 manu-draft.Rmd
personal edits after coauth revise

---
title: "Initial estuarine response to the nutrient-rich Piney Point discharge into Tampa Bay, Florida"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
    number_sections: false
always_allow_html: true
bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Andrew Altieri (andrew.altieri@essie.ufl.edu), Christine Angelini (christine.angelini@essie.ufl.edu), Maya C. Burke (mburke@tbep.org), Jing Chen (jchen15@usf.edu), Diana W. Chin (diwchin@gmail.com), Emma E. Dontis (edontis@co.pinellas.fl.us), Jayne Gardiner (jgardiner@ncf.edu), Jessica Lewis (jlewis@tbep.org), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J. Phlips (phlips@ufl.edu), Gary E. Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Edward T. Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Bob Weisberg (weisberg@usf.edu), Nijole Wellendorf (nijole.wellendorf@floridadep.org), Joe Whalen (jwhalen@tbep.org)"
urlcolor: blue
csl: marine-pollution-bulletin.csl
link-citations: true
---

@@

@@

Aging infrastructure and climate change stressors acting on wastewater containment facilities are global concerns for the environmental protection of surface waters.  From March 30th to April 9th, 2021, 814 million liters of mixed legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into Tampa Bay (Florida, USA).  An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.  In the period immediately after the discharge, elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay. Filamentous cyanobacteria blooms (*Dapis spp.*) peaked in June, followed by *Karenia brevis* blooms that peaked in July.  Reported fish kills tracked red tide concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.  By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the discharge from Piney Point are abnormal relative to historic conditions in Tampa Bay.

@@

Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11]. Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].  The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O), which is typically stored on-site in large earthen stacks (gypstacks) capable of storing millions of liters of wastewater.  Water quality in gypstacks can vary depending on processing method used at the mining facility, background geological characteristics of the region, and on-site practices for managing stormwater or other activities that can introduce additional materials to the holding ponds [@Henderson04;@Perez10].  In addition to elevated phosphorus concentrations, other nutrients, contaminants, and radionuclides may be present at values much higher than those of natural surface waters [@Burnett01;@Beck18b]. Many of these gypstacks no longer supporting active mining and aging infrastructure combined with climate change related stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain water on site. Environmental and human health risks associated with these stacks can occur through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].

Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite recent environmental recovery.  Successful restoration has been accomplished through ecosystem management paradigms that are based primarily on the control of nutrient pollutants from atmospheric deposition, stormwater, and wastewater.  Nitrogen inputs from external sources are well understood as drivers of algal blooms in coastal environments that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14]. Seagrasses in particular are valuable habitat-defining species that provide many ecosystem services, but are especially vulnerable to the impacts of nutrient pollution on water quality because of cascading effects of nitrogen, phytoplankton growth, water clarity, light limitation, and epiphytic growth that can eliminate seagrass from nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].  Historical gains in seagrass coverage in southwest Florida estuaries (e.g., Tampa Bay, Sarasota Bay, Charlotte Harbor) have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to surface waters [@Janicki96;@Greening16;@Tomasko18;@Tomasko20].  Together, these efforts have resulted in the long-term recovery of Tampa Bay and other Gulf Coast estuaries through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17]. The three contiguous National Estuary Programs of southwest Florida and their partners have been instrumental in coordinating efforts to address legacy pollutants and current threats to the long-term protection of estuarine resources.

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer to support agricultural production in other countries [@Henderson04].  By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, seventeen phosphogypsum stacks exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.  The potential environmental impacts of these facilities are considered a critical threat that could halt or reverse the decades of prior ecosystem improvement. The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].   Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.  Discharges from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.  Those discharges resulted in spatially-restricted, ecosystem responses including localized harmful algal blooms and increased macroalgal abundance [@Switzer11;@Garrett11].

@@

This paper provides an initial assessment of environmental conditions in Tampa Bay over six months following the recent release of legacy phosphate mining wastewater in April, 2021.  The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region. We focus on nitrogen inputs as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production.  A timeline of events in 2021 is provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.  The results from this paper provide examples of anticipated short-term responses to acute nutrient loadings from legacy mining facilities in the broader context of historical conditions that may influence system response to these events.

@@

Monitoring of the natural resources of Tampa Bay in response to the discharge from Piney Point began in April, 2021 and continued for six months through September. These data were collected through a coordinated effort under the guidance of a discharge plume simulation by a numerical circulation model run by Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science Ocean Circulation Lab.  The evolution of discharge from Piney Point was simulated using the Tampa Bay Coastal Ocean Model (TBCOM) nowcast/forecast system [@Chen18;@Chen19], with an embedded tracer module that included realistic discharge rates. Normalized tracer distributions were automatically updated each day, providing 1-day hindcasts and 3.5-day forecasts throughout the period of discharge and subsequent Tampa Bay distribution. The modeled plume evolution web product (Y. Liu, R.H. Weisberg, J. Chen, Y. Sun, personal communication) served as the principal guidance for coordinating the data collection during the event.

Monitoring agencies and local partners that collected data using standardized protocols included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St. Petersburg, Tampa Bay Estuary Program (TBEP), Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida. Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling (laboratory processing of discrete samples and *in situ* measurements), phytoplankton cell counts, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)). Details of the sampling methods are provided in supplement.

@@

Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above. These areas (Area 1: closest to Piney Point; Area 2: north of Piney Point; Area 3: south of Piney Point including northern Sarasota Bay) were identified based on anticipated impacts from expected plume patterns following the TBCOM simulations and other prominent bay boundaries relative to Piney Point (i.e., the main shipping channel in the bay, inflow boundaries, location of the Skyway Bridge at the mouth of Tampa Bay, and major bay segments used by TBEP for assessing programmatic goals). Observations at each monitoring station were averaged for each month across years from 2006 to 2020. This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 15,378 hectares (38,000 acres) [@Greening14;@Sherwood17].  For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest monitoring station that was sampled in response to Piney Point.  This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July). In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.  The average distance from a monitoring location in 2021 to the long-term sites was 1.6 km (see https://shiny.tbep.org/piney-point/ for a map of the matches).

The historical monitoring data were also used to model an expected seasonal pattern for water quality parameters from April to October in 2021.  This was done by estimating smoothed splines using Generalized Additive Models (GAMs) using data only from the "recovery" stage of Tampa Bay (2006 to 2020).  GAMs were used to model time series of water quality parameters as a function of a continuous value for year (i.e., decimal year) and as an integer value for day of year. The continuous year value was modeled with a thin plate regression spline and the day of year value was modelled with a cyclic spline [following similar methods @Murphy19].  The modeled results provided an estimate of the expected normal seasonal variation that takes into account a long-term annual trend.  Differences in the observed values sampled in the April to October time periods from the "forecasted" predictions of the baseline GAMs through 2021 provided an assessment of how the current data may have deviated from historical and normal seasonal variation.

Statistical assessments were evaluated only on total nitrogen (TN), chlorophyll-a (chl-a), and Secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred on different days during the five month period.  Spatial comparisons were based primarily on the three areas identified in Figure \@ref(fig:map)a. Variables with log-normal distributions were log$_10$-transformed (i.e., nutrients, chl-a) prior to analysis.  For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection dates. For censored data (i.e., above or below detection), Secchi observations that were visually identified on the bottom (`r sumb` of `r totb` observations in the FDEP data) were removed from analysis, whereas all other non-detects that were below laboratory standards of detection were evaluated with methods described below.

Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13]. Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons. An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.  For water quality variables, monthly averages from long-term monitoring data were subtracted from 2021 observations to account for normal seasonal variation not attributed to potential effects from Piney Point.  Similar corrections were not done for monthly comparisons of seagrass and macroalgae data because comparable long-term seasonal data do not exist. When appropriate, methods were used to accommodate measured concentrations in water quality variables that were below detection.  These included summary statistics (e.g., median, mean, and standard deviation) following estimates of the empirical cumulative distribution functions for each parameter using the Kaplan-Meier method for censored data [@Helsel05;@Lee20].

@@

Samples collected in the bay between April through September 2021 indicated that water quality conditions were unusual and outside of normal values expected for each month. A total of `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, TN, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, Secchi depth, and temperature (Table \@ref(tab:wqsumtab)). The percentage of observations outside of the normal range (mean +/- 1 standard deviation from long-term data) varied by location and parameter.  For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for Area 1 located closest to the discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 (to the north) and 3 (to the south), respectively. Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in Area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in Area 2 and `r tnabv[['3']]`\% in Area 3.  Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in Area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in Areas 2 and 3.  Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in Area 1, `r dosabv[['2']]`\% in Area 2).  Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in normal range, although initial time series showed much higher concentrations for ammonia in April near Area 1.  Ammonia concentrations were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab)). Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).  Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point (Area 1), Anna Maria Sound (Area 3), and the northern mouth of Tampa Bay (Area 3, Figure \@ref(fig:wqmap)).

Total nitrogen, chl-a, and Secchi depth followed a temporal progression in 2021 that was distinct from the long-term seasonal trend estimated from historical data (Figure \@ref(fig:wqgam)).  For Area 1, TN and chl-a concentrations were frequently above normal ranges during April.  Concentrations decreased slightly until June and July when values increased again above the seasonal expectation. Many Secchi observations in Area 1 were lower than normal in April and July.  Observations in Areas 2 and 3 were more often within the normal seasonal range, with some exceptions for TN and chl-a in Area 3 in  April, May, and July.  Statistical comparisons between months for TN, chl-a, and Secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgam). Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant (*p* < 0.05) for TN, chl-a, and Secchi depth for Area 1 and for TN and chl-a for Areas 2 and 3 (Table \@ref(tab:wqcmptab)). Further analsyis with multiple comparison tests generally showed that April/May were different from June/July depending on Area and parameter.  Observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K. brevis* abundances.

@@

```{r}
ntrnscnt <- length(unique(rstrndat$station))
ntrns <- english(ntrnscnt)
visdat <- rstrndat %>% 
  filter(date < maxdt) %>% 
  mutate(
    mo = month(date)
  ) %>% 
  group_by(station, mo) %>% 
  summarise(
    cnt = length(unique(date)), 
    .groups = 'drop'
  ) %>% 
  group_by(mo) %>% 
  summarise(
    cnt = mean(cnt)
  )
avevis <- round(mean(visdat$cnt), 1)

@@

# fo by month, area
fomo <- trnsum %>% 
  ungroup() %>% 
  mutate(
    foest = round(100 * foest, 0), 
    mo = month(date)
  )
fomoarea1 <- fomo %>% 
  filter(area == 'Area 1')
fomoarea3 <- fomo %>% 
  filter(area == 'Area 3')
fomoarea1ave <- fomoarea1 %>% 
  group_by(taxa) %>% 
  summarise(foest = round(mean(foest), 0)) %>% 
  deframe
fomoarea3ave <- fomoarea3 %>% 
  group_by(taxa) %>% 
  summarise(foest = round(mean(foest), 0)) %>% 
  deframe

>>>>

---
title: "Initial estuarine response to the nutrient-rich Piney Point discharge into Tampa Bay, Florida"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
    number_sections: false
always_allow_html: true
bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Andrew Altieri (andrew.altieri@essie.ufl.edu), Christine Angelini (christine.angelini@essie.ufl.edu), Maya C. Burke (mburke@tbep.org), Jing Chen (jchen15@usf.edu), Diana W. Chin (diwchin@gmail.com), Jayne Gardiner (jgardiner@ncf.edu), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J. Phlips (phlips@ufl.edu), Gary E. Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Edward T. Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Bob Weisberg (weisberg@usf.edu), Nijole Wellendorf (nijole.wellendorf@floridadep.org), Joe Whalen (jwhalen@tbep.org)"
urlcolor: blue
csl: marine-pollution-bulletin.csl
link-citations: true
---

@@

# seagrass transect counts
ntrnscnt <- length(unique(rstrndat$station))
ntrns <- english(ntrnscnt)

@@

Aging infrastructure and climate change stressors acting on wastewater containment facilities are global concerns for the environmental protection of surface waters.  From March 30th to April 9th, 2021, 814 million liters of mixed legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into Tampa Bay (Florida, USA).  An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.  In the period immediately after the discharge, elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay. Filamentous cyanobacteria blooms (*Dapis spp.*) peaked in June, followed by blooms of the red tide organism *Karenia brevis* that peaked in July.  Reported fish kills tracked *K. brevis* concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.  By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the discharge from Piney Point are abnormal relative to historic conditions in Tampa Bay.

@@

Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11]. Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].  The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O).  This waste is typically stored on-site in large earthen stacks (gypstacks) capable of holding millions of liters of process water.  Water quality in gypstacks can vary depending on processing method used at the mining facility, background geological characteristics of the region, and on-site practices for managing stormwater or other activities that can introduce additional materials to the holding ponds [@Henderson04;@Perez10].  In addition to elevated phosphorus concentrations, other nutrients, contaminants, and radionuclides may be present at values much higher than those of natural surface waters [@Burnett01;@Beck18b]. Many of these gypstacks no longer support active mining and aging infrastructure combined with climate change related stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain water on site. Environmental and human health risks associated with these stacks can occur through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].

Ongoing threats and challenges to protecting water quality of northern Gulf of Mexico estuaries persist despite recent environmental recovery.  Successful restoration has been accomplished through ecosystem management paradigms that are based primarily on the control of nutrient pollutants from atmospheric deposition, stormwater, and wastewater.  Nitrogen inputs from external sources are well understood as drivers of algal blooms in coastal environments that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14]. Seagrasses in particular are valuable habitat-defining species that provide many ecosystem services, but are especially vulnerable to the impacts of nutrient pollution on water quality because of cascading effects of nitrogen, phytoplankton growth, water clarity, light limitation, and epiphytic growth that can eliminate seagrass from nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].  Historical gains in seagrass coverage in southwest Florida estuaries (e.g., Tampa Bay, Sarasota Bay, Charlotte Harbor) have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to surface waters [@Janicki96;@Greening16;@Tomasko18;@Tomasko20].  Together, these efforts have resulted in the long-term recovery of Tampa Bay and other Gulf Coast estuaries through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17]. The three contiguous National Estuary Programs of southwest Florida and their partners have been instrumental in coordinating efforts to address legacy pollutants and current threats to the long-term protection of estuarine resources.

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer to support agricultural production in other countries [@Henderson04].  By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, seventeen phosphogypsum stacks exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.  The potential environmental impacts of these facilities are considered a critical threat that could halt or reverse the decades of prior ecosystem improvement. The Piney Point facility located in Palmetto, Florida is a large, remnant gypstack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [see supplement for a history of the facility, @Henderson04].   Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.  Discharges from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.  Those discharges resulted in spatially-restricted, ecosystem responses including localized harmful algal blooms and increased macroalgal abundance [@Switzer11;@Garrett11].

@@

This paper provides an initial assessment of environmental conditions in Tampa Bay over six months following the recent release of legacy phosphate mining wastewater from March 30th to April 9th, 2021.  The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region. We focus on nitrogen inputs as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production.  A timeline of events in 2021 is provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.  The results from this paper provide examples of anticipated short-term responses to acute nutrient loadings from legacy mining facilities in the broader context of historical conditions that may influence system response to these events.

@@

Monitoring of the natural resources of Tampa Bay in response to the discharge from Piney Point began in April, 2021 and continued for six months through September. These data were collected through a coordinated effort under the guidance of a discharge plume simulation by a numerical circulation model run by the Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science.  The evolution of discharge from Piney Point was simulated using the Tampa Bay Coastal Ocean Model (TBCOM) nowcast/forecast system [@Chen18;@Chen19], with an embedded tracer module that included realistic discharge rates. Normalized tracer distributions were automatically updated each day, providing 1-day hindcasts and 3.5-day forecasts throughout the period of discharge and subsequent Tampa Bay distribution. The modeled plume evolution web product (Y. Liu, R.H. Weisberg, J. Chen, Y. Sun, personal communication) served as the principal guidance for coordinating the data collection during the event.

Monitoring agencies and local partners that collected data using standardized protocols included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St. Petersburg, Tampa Bay Estuary Program (TBEP), Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida. Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling, phytoplankton identification, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).  In short, water quality parameters included discrete, laboratory-processed and *in situ* samples for total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a (ug/L), pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%).  Most samples were surface collections by boat, with sample frequency approximately biweekly for locations around Piney Point.  Phytoplankton samples included a mix of quantitative (cells/l) and qualitative (presence/absence) samples for major taxa at similar frequency and spatial distribution as the water quality samples. Seagrass and macroalgae sampling occurred approximately biweekly at `r ntrns` transects using a modified rapid assessment design following the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" method developed by a local citizen science group in cooperation with academic and federal partners.    Details of the sampling methods are provided in supplement.

@@

Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above. These areas (Area 1: closest to Piney Point; Area 2: north of Piney Point; Area 3: south of Piney Point including northern Sarasota Bay) were identified based on anticipated impacts from expected plume patterns following the TBCOM simulations and other prominent bay boundaries relative to Piney Point (i.e., the main shipping channel in the bay, inflow boundaries, location of the Skyway Bridge at the mouth of Tampa Bay, and major bay segments used by TBEP for assessing programmatic goals). Observations at each long-term monitoring station were averaged for each month across years from 2006 to 2020. This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 15,378 hectares [38,000 acres, @Greening14;@Sherwood17].  For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest water quality monitoring station that was sampled in response to Piney Point.  This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions near the mouth of Tampa Bay are not the same as those in the middle of the bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July). In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.  The average distance from a monitoring location in 2021 to the long-term sites was 1.6 km (see https://shiny.tbep.org/piney-point/ for a map of the matches).

The historical monitoring data were also used to model an expected seasonal pattern for water quality parameters from April to October in 2021.  This was done by estimating smoothed annual and seasonal splines with Generalized Additive Models (GAMs) using data only from the "recovery" stage of Tampa Bay (2006 to 2020).  GAMs were used to model time series of water quality parameters as a function of a continuous value for year (i.e., decimal year) and as an integer value for day of year. The continuous year value was modeled with a thin plate regression spline and the day of year value was modelled with a cyclic spline [following similar methods @Murphy19].  The modeled results provided an estimate of the expected normal seasonal variation that takes into account a long-term annual trend.  Differences in the observed values sampled in the April to October time periods from the "forecasted" predictions of the baseline GAMs through 2021 provided an assessment of how the current data may have deviated from historical and normal seasonal variation.

Statistical assessments were evaluated only on total nitrogen (TN), chlorophyll-a (chl-a), and Secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred on different days during the six month period.  Spatial comparisons were based primarily on the three areas identified in Figure \@ref(fig:map)a. Variables with log-normal distributions were log$_10$-transformed (i.e., nutrients, chl-a) prior to analysis.  For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection dates. For censored data (i.e., above or below detection), Secchi observations that were visually identified on the bottom (`r sumb` of `r totb` observations in the FDEP data) were removed from analysis, whereas all other non-detects that were below laboratory standards of detection were evaluated with methods described below.

Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13]. Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons. An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.  For water quality variables, monthly averages from long-term monitoring data were subtracted from 2021 observations to account for normal seasonal variation not attributed to potential effects from Piney Point.  Similar corrections were not done for monthly comparisons of seagrass and macroalgae data because comparable long-term seasonal data do not exist. Methods to accommodate measured concentrations of water quality variables that were below detection were used that included summary statistics (e.g., median, mean, and standard deviation) following estimates of the empirical cumulative distribution functions for each parameter using the Kaplan-Meier method for censored data [@Helsel05;@Lee20].

@@

Samples collected in the bay between April through September 2021 indicated that water quality conditions were unusual and outside of normal values expected for each month. A total of `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, TN, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, Secchi depth, and temperature (Table \@ref(tab:wqsumtab)). The percentage of observations outside of the normal range (mean +/- 1 standard deviation from long-term data) varied by location and parameter.  For chl-a, `r chlabv[['1']]`\% of the observations from April through September were above the normal range for Area 1 located closest to the discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 (to the north) and 3 (to the south), respectively. Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in Area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in Area 2 and `r tnabv[['3']]`\% in Area 3.  Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in Area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in Areas 2 and 3.  Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in Area 1, `r dosabv[['2']]`\% in Area 2).  Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in normal range, although initial time series showed much higher concentrations for ammonia in April near Area 1.  Ammonia concentrations were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab)). Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).  Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point (Area 1), Anna Maria Sound (Area 3), and the northern mouth of Tampa Bay (Area 3, Figure \@ref(fig:wqmap)).

Total nitrogen, chl-a, and Secchi depth followed a temporal progression in 2021 that was distinct from the long-term seasonal trend estimated from historical data (Figure \@ref(fig:wqgam)).  For Area 1, TN and chl-a concentrations were frequently above normal ranges during April.  Concentrations decreased slightly until June and July when values increased again above the seasonal expectation, coincident with the increase in *K. brevis* concentrations. Many Secchi observations in Area 1 were lower than normal in April and July.  Observations in Areas 2 and 3 were more often within the normal seasonal range, with some exceptions for TN and chl-a in Area 3 in  April, May, and July.  Statistical comparisons between months for seasonally-corrected observations of TN, chl-a, and Secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgam). Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant (*p* < 0.05) for TN, chl-a, and Secchi depth for Area 1 and for TN and chl-a for Areas 2 and 3 (Table \@ref(tab:wqcmptab)). Further analysis with multiple comparison tests generally showed that April/May were different from June/July depending on Area and parameter, such that observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K. brevis* abundances by mid-summer.

@@

```{r}
visdat <- rstrndat %>% 
  filter(date < maxdt) %>% 
  mutate(
    mo = month(date)
  ) %>% 
  group_by(station, mo) %>% 
  summarise(
    cnt = length(unique(date)), 
    .groups = 'drop'
  ) %>% 
  group_by(mo) %>% 
  summarise(
    cnt = mean(cnt)
  )
avevis <- round(mean(visdat$cnt), 1)

@@

# fo by month, area
fomo <- trnsum %>% 
  ungroup() %>% 
  mutate(
    foest = round(100 * foest, 0), 
    mo = month(date)
  )
fomobrn <- fomo %>% 
  filter(taxa == 'Brown') %>% 
  filter(foest > 0) %>% 
  pull(foest)
fomoarea1 <- fomo %>% 
  filter(area == 'Area 1')
fomoarea3 <- fomo %>% 
  filter(area == 'Area 3')
fomoarea1ave <- fomoarea1 %>% 
  group_by(taxa) %>% 
  summarise(foest = round(mean(foest), 0)) %>% 
  deframe
fomoarea3ave <- fomoarea3 %>% 
  group_by(taxa) %>% 
  summarise(foest = round(mean(foest), 0)) %>% 
  deframe
